---
title: "Longitudinal Census Data Template"
subtitle: "PEP Population + ACS Socioeconomics + Decennial Anchors + Inflation Adjustment"
author: "Dan"
date: today
format:
  html:
    toc: true
    toc-depth: 2
    number-sections: true
    embed-resources: true
  pdf:
    toc: true
    number-sections: true
execute:
  echo: false
  warning: false
  message: false
params:
  # =========================================================================
  # CHANGE THESE FOR A DIFFERENT CITY
  # =========================================================================
  city: "Schertz"
  state: "TX"
  # =========================================================================
  # Year ranges for data pulls
  years_pep: !expr 2010:2024
  years_acs: !expr 2011:2023
  # Inflation base year (for constant-dollar conversions)
  base_year: 2024
  # Cache directories (keeps API traffic low; add to .gitignore)
  cache_census_dir: "cache/census"
  cache_inflation_dir: "cache/inflation"
---

```{r}
#| label: setup
#| include: false

# -----------------------------------------------------------------------------
# Package verification
# -----------------------------------------------------------------------------
required_pkgs <- c(
"tidycensus", "tigris", "dplyr", "purrr", "tibble", "stringr", "readr",
"digest", "lubridate", "quantmod", "zoo", "rlang", "ggplot2",
"flextable"
)

missing <- required_pkgs[
!base::vapply(required_pkgs, requireNamespace, logical(1L), quietly = TRUE)
]

if (base::length(missing) > 0L) {
base::stop(
  "Install missing packages:\n",
  base::paste(missing, collapse = ", ")
)
}

# -----------------------------------------------------------------------------
# Source helper functions
# -----------------------------------------------------------------------------
base::source("R/census_helpers.R")

# -----------------------------------------------------------------------------
# Options
# -----------------------------------------------------------------------------
base::options(tigris_use_cache = TRUE)
```

```{r}
#| label: fallback-params
#| include: false

# Fallback for interactive use outside Quarto render
if (!base::exists("params")) {
params <- base::list(
  city               = "Schertz",
  state              = "TX",
  years_pep          = 2010:2024,
  years_acs          = 2011:2023,
  base_year          = 2024,
  cache_census_dir   = "cache/census",
  cache_inflation_dir = "cache/inflation"
)
}
```

# Overview

This template generates a unified longitudinal dataset for **`r params$city`, `r params$state`** combining:

- **PEP**: Annual population estimates (2010–2024)
- **ACS 5-Year**: Selected socioeconomic indicators
- **Decennial**: Census anchors (2000, 2010, 2020)
- **CPI-U**: Inflation adjustment factors for constant-dollar analysis

All API calls are cached to minimize traffic. Output uses a normalized schema: `source`, `series`, `year`, `estimate`, `moe`.


# PEP Annual Population

Population Estimates Program data for `r params$city`.

```{r}
#| label: pep-pull

pep <- ds_get_pep_population_place(
city      = params$city,
state     = params$state,
years     = params$years_pep,
cache_dir = params$cache_census_dir
)
```

```{r}
#| label: pep-table

pep_display <- pep |>
dplyr::arrange(year) |>
dplyr::mutate(
  Population = base::format(estimate, big.mark = ",", scientific = FALSE)
) |>
dplyr::select(Year = year, Source = source, Population)

ft_pep <- flextable::flextable(pep_display)
ft_pep <- ds_style_flextable(
ft_pep,
title = base::paste0("PEP Population: ", params$city, ", ", params$state)
)
ft_pep <- flextable::colformat_int(ft_pep, j = "Year", big.mark = "")

ft_pep

```


# ACS 5-Year Indicators

Selected American Community Survey variables. Modify `acs_vars` below to change which indicators are pulled.

```{r}
#| label: acs-vars

# -------------------------------------------------------------------------
# ACS VARIABLE CODES - MODIFY THIS SECTION FOR DIFFERENT INDICATORS
# -------------------------------------------------------------------------
# Format: friendly_name = "TABLE_VARIABLE"
# Common examples:
#   B19013_001 = Median household income
#   B25001_001 = Total housing units
#   B01003_001 = Total population (ACS estimate)
#   B25077_001 = Median home value
#   B23025_005 = Unemployed (civilian labor force)
# -------------------------------------------------------------------------

acs_vars <- c(
median_hh_income = "B19013_001",
housing_units    = "B25001_001",
population_acs   = "B01003_001"
)
```

```{r}
#| label: acs-pull

acs <- ds_get_acs_place(
city      = params$city,
state     = params$state,
years     = params$years_acs,
variables = base::unname(acs_vars),
survey    = "acs5",
cache_dir = params$cache_census_dir
)

# Map variable codes to friendly labels
var_labels <- stats::setNames(base::names(acs_vars), acs_vars)
acs$series_label <- var_labels[acs$series]
acs$series_label <- base::ifelse(
base::is.na(acs$series_label),
acs$series,
acs$series_label
)
```

```{r}
#| label: acs-table

acs_display <- acs |>
dplyr::arrange(series_label, year) |>
dplyr::mutate(
  estimate = base::format(
    base::round(estimate, 0),
    big.mark = ",",
    scientific = FALSE
  ),
  moe = base::ifelse(
    base::is.na(moe),
    "—",
    base::format(base::round(moe, 0), big.mark = ",")
  )
) |>
dplyr::select(
  Year     = year,
  Variable = series_label,
  Estimate = estimate,
  MOE      = moe
)

ft_acs <- flextable::flextable(acs_display)
ft_acs <- ds_style_flextable(
ft_acs,
title = base::paste0("ACS 5-Year Estimates: ", params$city, ", ", params$state)
)
ft_acs <- flextable::colformat_int(ft_acs, j = "Year", big.mark = "")

ft_acs
```


# Decennial Census Anchors

Official Census counts from 2000, 2010, and 2020.

```{r}
#| label: decennial-pull

# Variable codes differ by census year
# 2000/2010 SF1: P001001 (total population)
# 2020 PL: P1_001N (total population)

dec_2000 <- ds_get_decennial_place(
params$city, params$state,
year = 2000L, variable = "P001001", sumfile = "sf1",
cache_dir = params$cache_census_dir
)

dec_2010 <- ds_get_decennial_place(
params$city, params$state,
year = 2010L, variable = "P001001", sumfile = "sf1",
cache_dir = params$cache_census_dir
)

dec_2020 <- ds_get_decennial_place(
params$city, params$state,
year = 2020L, variable = "P1_001N", sumfile = "pl",
cache_dir = params$cache_census_dir
)

decennial <- dplyr::bind_rows(dec_2000, dec_2010, dec_2020)
decennial$series <- "population_decennial"
```

```{r}
#| label: decennial-table

dec_display <- decennial |>
dplyr::arrange(year) |>
dplyr::mutate(
  Population = base::format(estimate, big.mark = ",", scientific = FALSE)
) |>
dplyr::select(Year = year, Source = source, Population)

ft_dec <- flextable::flextable(dec_display)
ft_dec <- ds_style_flextable(
ft_dec,
title = base::paste0("Decennial Census: ", params$city, ", ", params$state)
)
ft_dec <- flextable::colformat_int(ft_dec, j = "Year", big.mark = "")

ft_dec
```


# Unified Longitudinal Dataset

Combined dataset with normalized schema across all sources.

```{r}
#| label: unify

pep_long <- pep
pep_long$series_label <- "population_pep"

acs_long <- acs
acs_long$series_label <- base::ifelse(
base::is.na(acs_long$series_label),
acs_long$series,
acs_long$series_label
)

dec_long <- decennial
dec_long$series_label <- "population_decennial"

longitudinal <- dplyr::bind_rows(pep_long, acs_long, dec_long)
longitudinal <- longitudinal |>
dplyr::select(source, series_label, year, estimate, moe) |>
dplyr::arrange(series_label, year)
```

```{r}
#| label: longitudinal-summary

summary_tbl <- longitudinal |>
dplyr::group_by(series_label) |>
dplyr::summarise(
  year_min = base::min(year, na.rm = TRUE),
  year_max = base::max(year, na.rm = TRUE),
  n_obs    = dplyr::n(),
  .groups  = "drop"
)

ft_summary <- flextable::flextable(summary_tbl)
ft_summary <- flextable::set_header_labels(
ft_summary,
series_label = "Series",
year_min     = "First Year",
year_max     = "Last Year",
n_obs        = "Observations"
)
ft_summary <- flextable::colformat_int(ft_summary, j = c("year_min", "year_max"), big.mark = "")
ft_summary <- ds_style_flextable(ft_summary, title = "Longitudinal Dataset Summary")

ft_summary
```


# Inflation Adjustment

CPI-U annual averages from FRED for converting nominal dollars to `r params$base_year` constant dollars.

```{r}
#| label: cpi-pull

year_min <- base::min(c(params$years_acs, params$years_pep), na.rm = TRUE)
year_max <- base::max(c(params$years_acs, params$years_pep), na.rm = TRUE)

cpi_tbl <- ds_get_cpi_annual(
start_year = year_min,
end_year   = year_max,
base_year  = params$base_year,
cache_dir  = params$cache_inflation_dir
)
```

```{r}
#| label: cpi-table

cpi_display <- cpi_tbl |>
dplyr::mutate(
  year = base::as.character(year),
  base_year = base::as.character(base_year),
  cpi_annual_avg = base::round(cpi_annual_avg, 2),
  factor_to_base = base::round(factor_to_base, 4)
) |>
dplyr::select(
  Year        = year,
  `CPI-U Avg` = cpi_annual_avg,
  `Base Year` = base_year,
  Factor      = factor_to_base
)

ft_cpi <- flextable::flextable(cpi_display)
ft_cpi <- ds_style_flextable(
  ft_cpi,
  title = base::paste0("CPI-U Inflation Factors (Base: ", params$base_year, ")")
)
ft_cpi
```


# Real Income Analysis

Median household income adjusted to `r params$base_year` dollars.
```{r}
#| label: income-real

income_nominal <- acs_long |>
dplyr::filter(series_label == "median_hh_income") |>
dplyr::select(source, series_label, year, estimate, moe)

if (base::nrow(income_nominal) > 0L) {
income_real <- ds_adjust_to_base_dollars(
  df        = income_nominal,
  value_col = "estimate",
  year_col  = "year",
  cpi_tbl   = cpi_tbl,
  out_col   = "estimate_real"
)
} else {
income_real <- tibble::tibble()
}
```

```{r}
#| label: income-table
#| eval: !expr nrow(income_real) > 0

income_display <- income_real |>
dplyr::arrange(year) |>
dplyr::mutate(
  Year          = base::as.character(year),
  Nominal       = base::paste0("$", base::format(base::round(estimate, 0), big.mark = ",")),
  Real          = base::paste0("$", base::format(base::round(estimate_real, 0), big.mark = ","))
) |>
dplyr::select(Year, Nominal, Real)

ft_income <- flextable::flextable(income_display)
ft_income <- ds_style_flextable(
  ft_income,
  title = base::paste0("Median Household Income (", params$base_year, " Dollars)")
)
ft_income
```


# Population Trend (XmR Chart)

Statistical process control chart for PEP population estimates.

```{r}
#| label: xmr-chart
#| fig-width: 8
#| fig-height: 5

pop_chart <- pep_long |>
dplyr::select(year, estimate) |>
dplyr::arrange(year)

ds_plot_xmr(
df        = pop_chart,
year_col  = "year",
value_col = "estimate",
title     = base::paste0("Population XmR: ", params$city, ", ", params$state)
)

```


# Export Data

```{r}
#| label: export

# base::dir.create("out", showWarnings = FALSE, recursive = TRUE)
# 
# readr::write_csv(
# longitudinal,
# base::file.path("out", "longitudinal_place_series.csv")
# )
# 
# readr::write_csv(
# cpi_tbl,
# base::file.path("out", "inflation_cpi_table.csv")
# )
# 
# if (base::nrow(income_real) > 0L) {
# readr::write_csv(
#   income_real,
#   base::file.path("out", "income_real_base_year.csv")
# )
# }

```

Data exported to `out/` directory:

- `longitudinal_place_series.csv` — All series combined
- `inflation_cpi_table.csv` — CPI factors
- `income_real_base_year.csv` — Inflation-adjusted income


# Using This Template for Other Cities
To analyze a different city, change only these parameters in the YAML header:

```yaml
params:
city: "Austin"
state: "TX"
```

Everything else adapts automatically. Clear your cache directory if switching cities:

```r
unlink("cache", recursive = TRUE)
```


## Notes

- **1990 Decennial**: Not supported by tidycensus for place-level data. Use NHGIS for earlier years.
- **ACS Years**: ACS 5-year estimates are labeled by end-year. First available: 2009 (covering 2005–2009).
- **PEP 2020+**: Uses Census Bureau Excel files since tidycensus API coverage ends at 2019.
